% ADL2Flowchart disjointness & region analysis report
% Compile: pdflatex disjointness_report.tex  (run twice for TOC/refs)
%          or: latexmk -pdf disjointness_report.tex

\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, fit, backgrounds}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=blue!60!black,
  urlcolor=blue!60!black
}

\definecolor{codebg}{RGB}{248,248,248}
\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{codebg},
  frame=single,
  framerule=0.4pt,
  breaklines=true,
  columns=fullflexible,
  showstringspaces=false
}

\title{%
  \textbf{Disjointness and Region Analysis in ADL2Flowchart}\\[0.4em]
  \large Progress report for collaborators
}
\author{ADL2Flowchart development team}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
This report summarises work on \emph{single-file} disjointness and overlap analysis
in the ADL2Flowchart compiler (\texttt{smash}), from the first experimental
\texttt{-r} pass through SMT-backed region proofs (Z3), constraint IR export, and
automated regression. The effort combined (i)~corpus hardening so that 68/68
example ADL files parse and analyse without crash, (ii)~heuristic object and
region disjointness, (iii)~a structured region constraint intermediate
representation, and (iv)~optional SMT encoding with proven overlap/disjoint
verdicts on a growing linear fragment. All changes are on the \texttt{main}
branch of the repository as of June 2026.
\end{abstract}

\tableofcontents
\newpage

%=============================================================================
\section{Motivation and goals}
%=============================================================================

Analysis Description Language (ADL) files describe event selections as named
\emph{regions} (signal regions, control regions, bins) built from cuts on
physics objects. For reinterpretation, combination studies, and systematic
validation, it is valuable to know whether two regions can share events:

\begin{itemize}[noitemsep]
  \item \textbf{Disjoint} --- no event can satisfy both regions' constraints.
  \item \textbf{Overlapping} --- at least one event can pass both (existence).
  \item \textbf{Unknown} --- the tool cannot decide with current extraction.
\end{itemize}

ADL2Flowchart already parsed ADL and produced Graphviz flowcharts; the
disjointness programme added \textbf{static analysis} on top of the existing
AST and symbol tables, without changing the core parse $\rightarrow$ typecheck
$\rightarrow$ DOT pipeline.

\paragraph{Scope.} Analysis is \textbf{single-file}: cross-file alias alignment
beyond \texttt{object\_aliases.txt} is not attempted. Conclusions are stated
relative to an \emph{extracted fragment} of each region's cuts; incomplete
extraction is surfaced via coverage metrics (Section~\ref{sec:coverage}).

%=============================================================================
\section{Timeline and development phases}
%=============================================================================

Table~\ref{tab:phases} maps the main delivery phases. Git commits on
\texttt{main} provide a finer-grained audit trail.

\begin{table}[ht]
  \centering
  \caption{Major phases of the disjointness implementation.}
  \label{tab:phases}
  \begin{tabular}{@{}clp{7.2cm}@{}}
    \toprule
    Phase & Label & Deliverables \\
    \midrule
    0 & Corpus hygiene & Parser/lexer/semantic fixes; 68/68 examples pass;
        object attribute baseline; \texttt{VERIFICATION\_REPORT.md} \\
    1 & Constraint IR & \texttt{gatherRegionConstraints()}, interval/tag/size
        extraction; overlap heuristics; JSON export \\
    2 & Z3 spike & Subprocess SMT-LIB; Delphes CMS-SUS-16-033 validation script \\
    3 & CLI integration & \texttt{-r} with auto-Z3; \texttt{--no-smt};
        \texttt{--json}; golden tests \\
    4 & CI \& docs & GitHub Actions; \texttt{make test-disjoint},
        \texttt{make test-corpus} \\
    v2 & SMT robustness & Index-safe keys; QF\_LIRA; angular encoding;
        unified report; coverage in JSON \\
    v3 & Logic fragment & OR disjunctions; ITE implications; coverage warnings \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Branch history.} Feature work landed on \texttt{disjoint\_dev}, was
merged into \texttt{main} (commit \texttt{0804db0}), then extended through
SMT v2/v3 (\texttt{8fa48a9}--\texttt{b6658b1}).

%=============================================================================
\section{Compiler pipeline and analysis architecture}
%=============================================================================

Figure~\ref{fig:pipeline} shows where new analysis hooks attach. The default
user-facing entry point remains \texttt{./smash <file.adl>}; region analysis
is opt-in via \texttt{-r}.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
    node distance=0.55cm and 0.4cm,
    box/.style={draw, rounded corners, minimum width=3.6cm, minimum height=0.75cm,
                align=center, font=\small},
    arr/.style={-{Stealth[length=2mm]}, thick}
  ]
    \node[box] (lex) {Flex lexer};
    \node[box, below=of lex] (parse) {Bison parser $\rightarrow$ AST};
    \node[box, below=of parse] (sem) {Semantic checks};
    \node[box, below=of sem] (dot) {AST \& flowchart DOT};
    \node[box, below=of dot, fill=blue!8] (obj) {Object disjointness};
    \node[box, below=of obj, fill=blue!8] (reg) {Region analysis IR + Z3};
    \node[box, below=of reg, fill=gray!10, dashed] (leg) {Legacy region report\\
      (\texttt{--legacy-region-report})};

    \draw[arr] (lex) -- (parse);
    \draw[arr] (parse) -- (sem);
    \draw[arr] (sem) -- (dot);
    \draw[arr] (dot) -- (obj);
    \draw[arr] (obj) -- (reg);
    \draw[arr, dashed] (reg) -- (leg);

    \begin{scope}[on background layer]
      \node[fit=(obj)(reg)(leg), draw=blue!40, rounded corners, inner sep=0.25cm,
            label={[font=\small]right:\texttt{-r}}] {};
    \end{scope}
  \end{tikzpicture}
  \caption{ADL2Flowchart pipeline with optional disjointness analysis (\texttt{-r}).}
  \label{fig:pipeline}
\end{figure}

\paragraph{Key modules.}
\begin{itemize}[noitemsep]
  \item \texttt{adl/semantic\_checks.cpp} --- constraint extraction, legacy
        \texttt{analyzeRegionDisjointness}, \texttt{analyzeObjectDisjointness},
        \texttt{gatherRegionConstraints}.
  \item \texttt{adl/region\_analysis.cpp} --- IR pairwise analysis, Z3 driver,
        JSON export, coverage warnings.
  \item \texttt{adl/object\_aliases.txt} --- canonical naming for take roots.
  \item \texttt{tests/golden/} --- small ADL fixtures with expected verdicts.
\end{itemize}

%=============================================================================
\section{Corpus hardening (Phase 0)}
%=============================================================================

Before disjointness could be trusted on real analyses, the example corpus had to
parse reliably.

\subsection{Baseline vs.\ current status}

\begin{table}[ht]
  \centering
  \caption{Example corpus verification (\texttt{examples/**/*.adl}, $N{=}68$).}
  \label{tab:corpus}
  \begin{tabular}{@{}lcc@{}}
    \toprule
    Metric & Before disjoint work & Current \\
    \midrule
    Parse success & 49 / 68 & \textbf{68 / 68} \\
    Post-parse crashes (exit 134) & 5 & \textbf{0} \\
    \texttt{./smash -r} without error & --- & \textbf{68 / 68} \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Representative fixes}

\begin{description}[style=nextline, font=\normalfont\ttfamily]
  \item[Lexer] Delphes-style IDs (\texttt{Delphes\_Photon}); path-like tokens;
        single-letter IDs for expressions such as \texttt{matchedCaloEnergy / p < 0.20}.
  \item[Grammar] Unary minus; chained ternaries; \texttt{print}/\texttt{save}/\texttt{counts};
        CutLang aliases; \texttt{object X : JET/COMB}; function-form \texttt{take}.
  \item[Semantics] Null-safe \texttt{binOpCheck}, \texttt{printITE}, \texttt{typeCheck};
        skip PRINT/SAVE in region declaration checks; dotted properties (\texttt{MET.pT}).
  \item[Stdlib] Extensions to \texttt{ext\_lib.txt} and \texttt{ext\_objs.txt}.
  \item[Stability] Fixed infinite recursion in \texttt{resolveParticleFamilies}
        (stack overflow on large \texttt{-r} runs).
\end{description}

Automated sweep: \texttt{make test-corpus} runs parse and \texttt{-r} on all 68 files.

%=============================================================================
\section{Object disjointness}
%=============================================================================

\texttt{analyzeObjectDisjointness} classifies pairs of user-defined \texttt{object}
declarations using take-lineage and \texttt{typeTable} particle families.

\paragraph{Verdicts.}
\begin{itemize}[noitemsep]
  \item \textbf{PROVEN DISJOINT} --- resolved families are incompatible (e.g.\
        electron vs.\ jet families with no shared ancestor).
  \item \textbf{POSSIBLY OVERLAPPING} --- shared lineage or unresolved family.
  \item \textbf{UNKNOWN} --- could not resolve particle family.
\end{itemize}

This pass is informational and runs automatically with \texttt{-r}. It
complements region analysis: disjoint objects often motivate disjoint region
cuts, but region-level proofs require the constraint IR below.

%=============================================================================
\section{Region constraint extraction}
%=============================================================================

\subsection{Supported cut forms}

The extractor walks region \texttt{select}/\texttt{reject}/\texttt{bin} conditions
(with inheritance) and builds a normalised constraint set per region.

\begin{table}[ht]
  \centering
  \caption{Constraint forms recognised in the IR (non-exhaustive).}
  \label{tab:extract}
  \begin{tabular}{@{}ll@{}}
    \toprule
    ADL surface syntax & IR key / notes \\
    \midrule
    \texttt{MET.pT > 200}, \texttt{pT(jets[0]) > 100} & Interval on scalar key \\
    \texttt{size(bjets) >= 2} & \texttt{size(canonical\_object)} \\
    \texttt{BTag == 1}, tags & Discrete 0/1 \\
    \texttt{dPhi(jets[0], MHT) > 0.5} & Angular interval key \\
    \texttt{reject} cuts & Complemented interval \\
    \texttt{define} used in cut & Inlined where referenced \\
    Inherited region name & Constraints merged from parent region \\
    \texttt{A || B} & OR clause (v3) \\
    \texttt{cond ? cut : ALL} & Implication clause (v3) \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Canonical keys and indexing}

A recurring bug class was \emph{key collapse}: \texttt{jets[0].pt} and
\texttt{jets[1].pt} were merged because object lineage logic ignored bracket
indices. \textbf{v2} fixes:

\begin{itemize}[noitemsep]
  \item \texttt{canonicalConstraintKey} preserves \texttt{[i]} before property suffixes.
  \item \texttt{constraintKeysRelated} refuses to relate keys with different index
        suffixes.
  \item Union-find in SMT only merges alias-equivalent keys (e.g.\ \texttt{JET[0].pt}
        with \texttt{jets[0].pt}), not different jet indices.
\end{itemize}

\paragraph{Important distinction.} Cuts on \texttt{jets[0]} and \texttt{jets[1]} are
\emph{independent variables} in the fragment: disjoint intervals on different indices
do \textbf{not} imply region disjointness (the same event can have a hard leading jet
and a soft subleading jet).

%=============================================================================
\section{Region pairwise analysis and verdicts}
%=============================================================================

For each unordered pair of regions $(R_1, R_2)$ the tool reports a relation:

\begin{table}[ht]
  \centering
  \caption{Region pairwise verdicts (IR + optional Z3).}
  \label{tab:verdicts}
  \begin{tabular}{@{}p{3.2cm}p{9.5cm}@{}}
    \toprule
    Verdict & Meaning \\
    \midrule
    PROVEN DISJOINT &
    Heuristic: disjoint intervals on a shared canonical key; or Z3 UNSAT on
    $R_1 \wedge R_2$ in the encoded fragment. UNSAT is sound for the fragment. \\
    PROVEN OVERLAPPING &
    Z3 SAT on $R_1 \wedge R_2$ with at least one \emph{shared constraint dimension};
    witness model printed. Existential in the fragment, not simulation yield. \\
    POSSIBLY OVERLAPPING &
    Heuristic interval compatibility; or SAT without shared dimension (independent cuts). \\
    UNKNOWN &
    Missing extraction, timeout, or inconclusive Z3 response. \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Legacy report.} The original verbose \texttt{analyzeRegionDisjointness}
printer remains available as \texttt{--legacy-region-report}. Default \texttt{-r}
output uses the unified \texttt{region\_analysis} summary only.

%=============================================================================
\section{SMT integration (Z3)}
%=============================================================================

\subsection{Design choices}

\begin{itemize}[noitemsep]
  \item \textbf{Subprocess, not libz3} --- SMT-LIB scripts via \texttt{z3} on PATH;
        avoids link-time dependency; CI installs \texttt{z3} via apt.
  \item \textbf{Logic} --- \texttt{QF\_LIRA} when any \texttt{size(...)} constraint
        uses \texttt{Int}; otherwise \texttt{QF\_LRA} for scalar intervals.
  \item \textbf{Auto-enable} --- \texttt{-r} runs Z3 when installed; \texttt{--no-smt}
        disables it.
  \item \textbf{Timeout} --- 15\,s per pair check (configurable in source).
\end{itemize}

\subsection{Encoding sketch}

For each region, constraints become a conjunction of:
\begin{itemize}[noitemsep]
  \item interval assertions on declared variables (one per canonical key);
  \item \textbf{OR} groups: \texttt{(assert (or (and ...) (and ...)))};
  \item \textbf{ITE}: \texttt{(assert (=> guard then))} and, if else $\neq$
        \texttt{ALL}, \texttt{(assert (=> (not guard) else))}.
\end{itemize}

Pairwise check: assert region~1 formula $\wedge$ region~2 formula, \texttt{check-sat};
optional \texttt{get-model} for overlap witnesses.

\subsection{Soundness caveats (present clearly to collaborators)}

\begin{enumerate}[noitemsep]
  \item Verdicts are over the \textbf{extracted fragment}, not full ADL semantics.
  \item BDTs, arbitrary functions, and deeply nested logic may be omitted.
  \item One SMT variable per canonical key --- no per-event multiplicities beyond
        \texttt{size()}.
  \item PROVEN OVERLAPPING = $\exists$ model satisfying linear constraints, not
        ``non-zero efficiency'' in simulation.
\end{enumerate}

%=============================================================================
\section{Fragment coverage and warnings}
\label{sec:coverage}
%=============================================================================

To avoid over-interpreting weak verdicts, \textbf{v3} reports coverage per region:

\begin{itemize}[noitemsep]
  \item \texttt{SMT-encodable $N$ / $M$ atoms} --- counts flat, OR, and ITE atoms.
  \item \texttt{selects encoded $K$ / $S$} --- fraction of select statements from
        which at least one constraint structure was extracted.
  \item \textbf{Warnings} if either ratio falls below \textbf{50\%}.
\end{itemize}

JSON export (\texttt{--json}) includes \texttt{fragment\_coverage} per region and a
top-level \texttt{coverage\_warnings} array for tooling and regression.

%=============================================================================
\section{Command-line interface}
%=============================================================================

\begin{lstlisting}
make
./smash file.adl                          # parse + DOT (unchanged)
./smash -r file.adl                       # object + region IR/SMT analysis
./smash -r --no-smt file.adl              # heuristics only
./smash -r --legacy-region-report file.adl
./smash -r --json analysis.json file.adl
make test-disjoint                        # golden fixtures
make test-corpus                          # 68 example files
./scripts/phase2_z3_spike.sh              # Delphes 033 SMT smoke test
\end{lstlisting}

%=============================================================================
\section{Testing and continuous integration}
%=============================================================================

\subsection{Golden fixtures (\texttt{tests/golden/})}

\begin{table}[ht]
  \centering
  \caption{Golden ADL tests and what they exercise.}
  \label{tab:golden}
  \small
  \begin{tabular}{@{}ll@{}}
    \toprule
    File & Intent \\
    \midrule
    \texttt{disjoint\_pt.adl} & Disjoint $p_\mathrm{T}$ intervals \\
    \texttt{disjoint\_jet\_index.adl} & Same-index disjoint intervals \\
    \texttt{independent\_jet\_index.adl} & Different indices may overlap \\
    \texttt{overlap\_met.adl} & Overlapping \texttt{MET.pT} intervals \\
    \texttt{size\_bjets.adl} & \texttt{size(bjets)} subset overlap (SMT) \\
    \texttt{or\_met.adl} & OR disjunction encoding \\
    \texttt{ite\_conditional\_dphi.adl} & ITE + angular cut (Delphes-style) \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{CI (\texttt{.github/workflows/ci.yml})}

On push/PR to \texttt{main}: install flex, bison, clang, graphviz, \textbf{z3};
\texttt{make}; \texttt{make test-disjoint}; \texttt{make test-corpus};
phase-2 Z3 spike on \texttt{CMS-SUS-16-033\_Delphes.adl}.

%=============================================================================
\section{Validation highlights}
%=============================================================================

\begin{itemize}[noitemsep]
  \item \textbf{68/68} example ADL files: parse and \texttt{-r} complete without crash.
  \item \textbf{Delphes 033}: dozens of \texttt{PROVEN OVERLAPPING [SMT]} pairs among
        signal regions (expected for nested $p_\mathrm{T}$/multiplicity ladders).
  \item \textbf{CMS-SUS-16-032}: structured BTag recognition; large disjoint pair count
        in legacy analysis after tag-key fix.
  \item All golden tests pass with Z3 4.8.x when installed.
\end{itemize}

%=============================================================================
\section{Known limitations and future work}
%=============================================================================

\begin{table}[ht]
  \centering
  \caption{Open items (not yet implemented).}
  \label{tab:future}
  \begin{tabular}{@{}p{4.5cm}p{8.5cm}@{}}
    \toprule
    Topic & Notes \\
    \midrule
    Cross-file analysis & Align region/object names across multiple ADL cards. \\
    Subset detection & Heuristic ``$R_2 \subseteq R_1$'' (deferred; easy false positives). \\
    Richer SMT & Nested OR/ITE flattening; per-jet indexed theories; BDT bounds. \\
    Grammar cleanup & 57 shift/reduce + 30 reduce/reduce Bison conflicts (pre-existing). \\
    Simulation link & Map SMT witnesses to CutLang execution / efficiencies. \\
    \bottomrule
  \end{tabular}
\end{table}

%=============================================================================
\section{Summary for collaborators}
%=============================================================================

We extended ADL2Flowchart from a visualisation-focused compiler to a tool that
can \textbf{audit region relationships} on real CMS/ATLAS-style ADL cards:

\begin{enumerate}[noitemsep]
  \item The \textbf{example corpus is production-stable} for parse and analysis.
  \item \textbf{Object and region analyses} give fast, explainable heuristics.
  \item A \textbf{constraint IR + JSON export} supports external tooling.
  \item \textbf{Z3} supplies machine-checked \emph{proven disjoint} (UNSAT) and
        \emph{proven overlapping} (SAT + witness) results on a documented linear
        fragment, including OR and Delphes-style ITE cuts.
  \item \textbf{Coverage warnings} flag when conclusions may be incomplete.
\end{enumerate}

For hands-on exploration, start with:
\begin{lstlisting}
./smash -r examples/CMS/CMS-SUS-16-033_Delphes.adl
./smash -r --json /tmp/out.json tests/golden/ite_conditional_dphi.adl
\end{lstlisting}

Further reading in the repository: \texttt{REGION\_ANALYSIS.md},
\texttt{VERIFICATION\_REPORT.md}, \texttt{docs/SMT\_ROBUSTNESS\_PLAN.md}.

%=============================================================================
\appendix
\section{Repository file index}
%=============================================================================

\begin{longtable}{@{}p{5.5cm}p{8.5cm}@{}}
\toprule
Path & Role \\
\midrule
\endhead
\texttt{adl/semantic\_checks.cpp} & Extraction, object/legacy region analysis \\
\texttt{adl/region\_analysis.cpp} & IR pairwise analysis, Z3, JSON, warnings \\
\texttt{adl/region\_analysis.hpp} & Public analysis types and API \\
\texttt{adl/object\_aliases.txt} & Take-root canonicalisation \\
\texttt{adl/main.cpp} & CLI flags (\texttt{-r}, \texttt{--json}, etc.) \\
\texttt{scripts/run\_golden\_tests.sh} & Regression driver \\
\texttt{scripts/validate\_corpus.sh} & Full example sweep \\
\texttt{scripts/phase2\_z3\_spike.sh} & Delphes SMT smoke test \\
\texttt{.github/workflows/ci.yml} & Continuous integration \\
\texttt{tests/golden/*.adl} & Minimal ADL regression fixtures \\
\bottomrule
\end{longtable}

\section{Suggested talking points for a presentation}
\begin{itemize}
  \item Show before/after: 49 $\rightarrow$ 68 parse success; zero crashes on \texttt{-r}.
  \item Live demo: \texttt{disjoint\_pt.adl} $\rightarrow$ PROVEN DISJOINT; \texttt{size\_bjets.adl}
        $\rightarrow$ PROVEN OVERLAPPING with witness.
  \item Explain \textbf{fragment} honestly: SMT proves properties of the encoded cuts,
        not the full physics simulation.
  \item Position Z3 as optional but recommended (\texttt{apt install z3}).
\end{itemize}

\end{document}